---
title: "Assignment 4 - Task 1"
author: "Meet Muchhala"
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

# 

```{r}
library(tidyverse)
library(here)
library(ggfortify)

```

## Initializing and setting up data

```{r}

## Loading the dataset

world_env_vars <- read_csv(here('data', 'world_env_vars.csv'))

## Dropping NAs due to PCA analysis and assigning factor levels to accessibility column
world_env_vars_clean <- world_env_vars %>%
  janitor::clean_names() %>%
  drop_na() %>% 
  mutate(accessibility_to_cities =
                     case_when(accessibility_to_cities <= 100 ~ "Very High", 
                              accessibility_to_cities <= 500 ~ "High",
                              accessibility_to_cities <= 1000 ~ "Medium",
                              accessibility_to_cities <= 11000 ~ "Low")) %>% 
  mutate(accessibility_to_cities = factor(accessibility_to_cities, levels = c('Low', 'Medium', 'High', 'Very High')))


glimpse(world_env_vars_clean)
names(world_env_vars_clean)
```

## PCA Analysis 

In this PCA analysis, the relationship between accessibility to cities and other wind variables in a country would be explored through the identification of principal components that capture the underlying patterns and relationships in the data. The inclusion of accessibility to cities as a variable in the PCA can help uncover how it relates to other factors influencing wind energy installations and identify potential patterns or clusters in the data across the world. 

```{r}

## Selecing relevant variables in case of building PCA analysis for wind energy installations and development country-wise

env_var_pca <- world_env_vars_clean %>% 
  select(wind, elevation, aspect, slope, rain_mean_annual, 
         temp_mean_annual, temp_annual_range, cloudiness) %>% 
  select(where(is.numeric)) %>%
  prcomp(scale = TRUE)

env_var_pca$rotation

```

## Bi-plot 

```{r}

autoplot(env_var_pca,
     	data = world_env_vars_clean,
     	loadings = TRUE,
     	colour = 'accessibility_to_cities',
     	loadings.label = TRUE,
     	loadings.colour = "black",
     	loadings.label.colour = "black",
     	loadings.label.vjust = -0.8,
     	loadings.label.hjust = 0.1
     	) +
  ggtitle("Accessibilities to cities globally for Wind Power - PCA") +
  guides(color = guide_legend(title = "Accessibility to Cities")) +
  scale_color_manual(values = c('red', 'orange', 'yellowgreen', 'darkgreen')) +
  theme_minimal()
```

* Mean annual rainfall and Annual temprature are negatively correlated. These two variables also have fairly long arrows, indicating larger variance.
* Mean annual rainfall and cloudiness are the most closely correlated; more rain is correlated with more 
cloudiness.
* Aspect has the shortest arrow, indicating less variance. 
* There may be some clustering of the "Very High" accessibility observations, as seen around the Annual temperature and wind arrow. 


## Screeplot

```{r}
sd_vec <- env_var_pca$sdev
var_vec <- sd_vec^2 ### standard deviation is sqrt of variance
pc_names <- colnames(env_var_pca$rotation)
pct_expl_df <- data.frame(v = var_vec,
                          pct_v = var_vec / sum(var_vec),
                          pc = fct_inorder(pc_names)) %>%
  mutate(pct_lbl = paste0(round(pct_v*100, 1), '%'))

ggplot(pct_expl_df, aes(x = pc, y = v)) +
  geom_col(color = "tomato4", fill = "turquoise") +
  ggtitle("Accessibilities to cities globally for Wind Power - Screeplot")+
  geom_text(aes(label = pct_lbl), vjust = 0, nudge_y = .002) +
  labs(x = 'Dimensions', y = 'Percentage of explained variances')+
  theme_minimal()
```

* The screeplot indicates four PCAs explain more than 80% of the variance. This may be a good cutoff point, beyond which additional components may not be as useful.